# Data cleaning and transformation
```{r,echo=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(vcd)
```
## Data Merging
Two datasets are provided regarding two distinct subjects: Mathematics (mat) and Portuguese language (por). To combine these two datasets, we spcified the following columns to be used for merging because these columns are basic information of students: \
```{r, echo = FALSE}
Description = readr::read_csv(here::here("./data/clean/description.csv"))
knitr::kable(Description[c(1:13,20,22), ],
             row.names = F,font_size = 12)
```

## Data Cleaning
After carefully inspecting each variable, we detected that there are outliers in `Age` and abnormal values in Math and Portuguese grades. \
Let's firstly take a look at the boxplot of `age`.\

```{r,echo=FALSE, fig.width = 7, fig.height = 3}
#Data Merging
mat=read.table("./data/raw/student-mat.csv",sep=";",header=TRUE)
por=read.table("./data/raw/student-por.csv",sep=";",header=TRUE)
df=merge(mat,por,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet","guardian","traveltime"))
colnames(df)[colnames(df) == 'G1.x'] = 'Math_G1'
colnames(df)[colnames(df) == 'G2.x'] = 'Math_G2'
colnames(df)[colnames(df) == 'G3.x'] = 'Math_G3'
colnames(df)[colnames(df) == 'G1.y'] = 'Portuguese_G1'
colnames(df)[colnames(df) == 'G2.y'] = 'Portuguese_G2'
colnames(df)[colnames(df) == 'G3.y'] = 'Portuguese_G3'
#Boxplot of age
ggplot(df,aes(x=factor(0),y=age))+geom_boxplot(fill='lightblue')+theme(axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank())+xlab("")+coord_flip()+theme_bw() +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

Given that most students leave secondary school at 16 or 17. There may be some exceptions so that age 19 in this dataset seems reasonable. However, we tend to believe that age 20 and 22 are outliers so it is better to remove data which include age = 20 or age = 22. \
Next, let's move to Math and Portuguese grades in different period. \
\
```{r,echo=FALSE}
math_long = df %>% pivot_longer(c(Math_G1,Math_G2,Math_G3), names_to = 'Period', values_to = 'Grade')
plot_math = ggplot(math_long,aes(x = Grade)) +
             geom_bar(fill = 'lightblue', color = 'black') +
             facet_wrap(~Period) +
             ggtitle('Histogram of Math Grades faceted by Period') +
             theme_bw() +
             theme(plot.title = element_text(hjust = 0.5),
                   panel.grid.major.x = element_blank(),
                   panel.grid.minor.x = element_blank())

portuguese_long = df %>% pivot_longer(c(Portuguese_G1,Portuguese_G2,Portuguese_G3), names_to = 'Period', values_to = 'Grade')
plot_portuguese = ggplot(portuguese_long,aes(x = Grade)) +
             geom_bar(fill = 'lightblue', color = 'black') +
             facet_wrap(~Period) +
             ggtitle('Histogram of Portuguese Grades faceted by Period') +
             theme_bw() +
             theme(plot.title = element_text(hjust = 0.5),
                   panel.grid.major.x = element_blank(),
                   panel.grid.minor.x = element_blank())
grid.arrange(plot_math, plot_portuguese, nrow = 2)
```

There are grades that equal to 0. Usually, a student with bad performance on a course may result in very low grade, but not 0. A grade of 0 usually appears when a student missed the exam or he/she cheated in exam. However, absence from exam or cheating are not part of factors we are studying. In this case, we remove all data with grade of 0. 

```{r}
# Data cleaning
df = df[df$age<20 & df$Math_G2 != 0 & df$Math_G3 != 0 & df$Portuguese_G1 != 0 & df$Portuguese_G3!=0,]
df = df %>% mutate(Por_Letter_Grade_G1 = case_when(Portuguese_G1 >= 18 & Portuguese_G1 <= 20 ~ 'A',
                                        Portuguese_G1 >= 14 & Portuguese_G1 < 18 ~ 'B',
                                        Portuguese_G1 >= 10 & Portuguese_G1 < 14 ~ 'C',
                                        Portuguese_G1 >= 4 & Portuguese_G1 < 10 ~ 'D',
                                        Portuguese_G1 >= 0 & Portuguese_G1 < 4 ~ 'F'),
  Por_Letter_Grade_G2 = case_when(Portuguese_G2 >= 18 & Portuguese_G2 <= 20 ~ 'A',
                                        Portuguese_G2 >= 14 & Portuguese_G2 < 18 ~ 'B',
                                        Portuguese_G2 >= 10 & Portuguese_G2 < 14 ~ 'C',
                                        Portuguese_G2 >= 4 & Portuguese_G2 < 10 ~ 'D',
                                        Portuguese_G2 >= 0 & Portuguese_G2 < 4 ~ 'F'),
  Por_Letter_Grade_G3 = case_when(Portuguese_G3 >= 18 & Portuguese_G3 <= 20 ~ 'A',
                                        Portuguese_G3 >= 14 & Portuguese_G3 < 18 ~ 'B',
                                        Portuguese_G3 >= 10 & Portuguese_G3 < 14 ~ 'C',
                                        Portuguese_G3 >= 4 & Portuguese_G3 < 10 ~ 'D',
                                        Portuguese_G3 >= 0 & Portuguese_G3 < 4 ~ 'F'),
  Math_Letter_Grade_G1 = case_when(Math_G1 >= 18 & Math_G1 <= 20 ~ 'A',
                                        Math_G1 >= 14 & Math_G1 < 18 ~ 'B',
                                        Math_G1 >= 10 & Math_G1 < 14 ~ 'C',
                                        Math_G1 >= 4 & Math_G1 < 10 ~ 'D',
                                        Math_G1 >= 0 & Math_G1 < 4 ~ 'F' ),
  Math_Letter_Grade_G2 = case_when(Math_G2 >= 18 & Math_G2 <= 20 ~ 'A',
                                        Math_G2 >= 14 & Math_G2 < 18 ~ 'B',
                                        Math_G2 >= 10 & Math_G2 < 14 ~ 'C',
                                        Math_G2 >= 4 & Math_G2 < 10 ~ 'D',
                                        Math_G2 >= 0 & Math_G2 < 4 ~ 'F' ),
  Math_Letter_Grade_G3 = case_when(Math_G3 >= 18 & Math_G3 <= 20 ~ 'A',
                                        Math_G3 >= 14 & Math_G3 < 18 ~ 'B',
                                        Math_G3 >= 10 & Math_G3 < 14 ~ 'C',
                                        Math_G3 >= 4 & Math_G3 < 10 ~ 'D',
                                        Math_G3 >= 0 & Math_G3 < 4 ~ 'F' ),
         absences.x = case_when(absences.x == 0 ~ 0,
                                absences.x >0 & absences.x <= 2 ~ 1,
                                absences.x >2 & absences.x <= 4 ~ 2,
                                absences.x >4 & absences.x <= 8 ~ 3,
                                absences.x >8 ~ 4),
         absences.y = case_when(absences.y == 0 ~ 0,
                                absences.y >0 & absences.y <= 2 ~ 1,
                                absences.y >2 & absences.y <= 4 ~ 2,
                                absences.y >4 ~ 3))
write.csv(df,'./data/clean/df.csv',row.names = FALSE)
```

## Variable Transformation

### Absences

We convert `absences` variable(numeric from 0 to 93) into categorical variable as follow: \
If `absences` = 0, then convert it to 0.\
If `absences` > 0 and `absences` <= first quartile, then convert it to 1. \
If `absences` > first quartile and `absences` <= second quartile, then convert it to 2.\
If `absences` > second quartile and `absences` <= thrid quartile, then convert it to 3. \
If `absences` > thrid quartile, then convert it to 4. \

### Grades

We convert all Math and Portuguese grades into categorical variables based on the following [grading system](https://en.wikipedia.org/wiki/Grading_systems_by_country#Portugal) for Portuguese high school: \
```{r,echo=FALSE}
gradingscale = readr::read_csv(here::here("./data/clean/grading_scale.csv"))
knitr::kable(gradingscale,
             row.names = F,font_size = 12)
```
