# Data cleaning and transformation
```{r,echo=FALSE}
library(ggplot2)
library(dplyr)
library(gridExtra)
library(tidyverse)
library(vcd)
```
## Data Merging
Two datasets are provided regarding two distinct subjects: Mathematics (mat) and Portuguese language (por). To combine these two datasets, we spcified the following columns to be used for merging because these columns are basic information of students: \
```{r, echo = FALSE}
Description = readr::read_csv(here::here("./data/Description.csv"))
knitr::kable(Description[c(1:12,20,22), ],
             row.names = F,font_size = 12)
```

## Data Cleaning
After inspecting each variables carefully, we detected that there were outliers in `Age` and abnormal values in Math and Portuguese grades. \
Let's first take a look at the boxplot of age.\
```{r,echo=FALSE}
mat=read.table("./data/student-mat.csv",sep=";",header=TRUE)
por=read.table("./data/student-por.csv",sep=";",header=TRUE)
df=merge(mat,por,by=c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet","guardian"))
colnames(df)[colnames(df) == 'G1.x'] = 'Math_G1'
colnames(df)[colnames(df) == 'G2.x'] = 'Math_G2'
colnames(df)[colnames(df) == 'G3.x'] = 'Math_G3'
colnames(df)[colnames(df) == 'G1.y'] = 'Portuguese_G1'
colnames(df)[colnames(df) == 'G2.y'] = 'Portuguese_G2'
colnames(df)[colnames(df) == 'G3.y'] = 'Portuguese_G3'
ggplot(df,aes(x=factor(0),y=age))+geom_boxplot(fill='lightblue')+theme(axis.title.y=element_blank(),axis.text.y=element_blank(),axis.ticks.y=element_blank())+coord_flip()
```

Most students leave secondary school at 16 or 17. There may be some exceptions and age 19 in this dataset seems reasonable. However, we tend to believe that age 20 and 22 are outliers so it is better to remove data which includes age = 20 or age = 22. \
Next, let's move to Math and Portuguese grades in different period. \
```{r,echo=FALSE}
math_long = df %>% pivot_longer(c(Math_G1,Math_G2,Math_G3), names_to = 'Period', values_to = 'Grade')
plot_math = ggplot(math_long,aes(x = Grade)) +
             geom_bar(fill = 'lightblue', color = 'black') +
             facet_wrap(~Period) +
             ggtitle('Histogram of Math Grades faceting by Period')

portuguese_long = df %>% pivot_longer(c(Portuguese_G1,Portuguese_G2,Portuguese_G3), names_to = 'Period', values_to = 'Grade')
plot_portuguese = ggplot(portuguese_long,aes(x = Grade)) +
             geom_bar(fill = 'lightblue', color = 'black') +
             facet_wrap(~Period) +
             ggtitle('Histogram of Portuguese Grades faceting by Period')
grid.arrange(plot_math, plot_portuguese, nrow = 2)
```

We can see there are grades that equal to 0. Usually, a student with bad performance on a course may result in very low grade, but not 0. A grade of 0 usually appears when the student misses the exam or he/she cheats in exam. However, absence from exam or cheating are not part of factors we are studying. In this case, we remove all data with grade of 0. 

## Variable Transformation
1. Convert `absences` variable(numeric from 0 to 93) into categorical variable as follow: \
If `absences` = 0, then convert it to 0.\
If `absences` > 0 and `absences` <= first quartile, then convert it to 1. \
If `absences` > first quartile and `absences` <= second quartile, then convert it to 2.\
If `absences` > second quartile and `absences` <= thrid quartile, then convert it to 3. \
If `absences` > thrid quartile, then convert it to 4. \

2. Convert all Math and Portuguese grades into categorical variables based on the following grading system for Portuguese school: \
```{r,echo=FALSE}
gradingscale = readr::read_csv(here::here("./data/Grading_Scale.csv"))
knitr::kable(gradingscale,
             row.names = F,font_size = 12)
```
The grading system can be found [here](https://en.wikipedia.org/wiki/Grading_systems_by_country#Portugal)