---
output: html_document
editor_options: 
chunk_output_type: console
---
# Results

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(vcd)
library(HH)
library(here)
library(readr)
```

## Students & Schools Analysis

In this section, we will analyze students' non-academic data and in two schools.

### Age

```{r}
stu <- readr::read_csv(here::here("./data/stu-mat.csv")) %>% 
  mutate(sex = ifelse(sex == "F","Female", "Male"),
         address = ifelse(address == "U","Urban ", "Rural"))
ggplot(stu, aes(factor(age))) +
  geom_bar(color = "black", 
           fill = "lightblue") +
  facet_wrap(~school) +
  xlab("Age") +
  ylab("Frequency") +
  ggtitle("Bar chart for the age by schools") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```
The bar chart plot below displays students' age distribution in the two schools. One of obvious property is that **GP** has much **more** students than **MS**. Precisely, **GP** has a total of `r sum(ifelse(stu$school=="GP",1,0))` students of which accounts for `r round(sum(ifelse(stu$school=="GP",1,0))/dim(stu)[1],4)*100`% of the dataset. Possibly due to the small number of students, age range is more narrow in **MS** (17-21) than that in **GP** (15-22).

In order to compare the age distribution effectively, we calculate relative frequency plot below.

```{r}
stu %>% 
  count(school, age) %>% 
  ungroup() %>% 
  group_by(school) %>% 
  mutate(per = n / sum(n)) %>% 
  dplyr::select(-n) %>% 
  ungroup() %>% 
  add_row(school = c("GP","MS","MS","MS"), 
          age = c(21, 15, 16, 22), 
          per = c(0, 0, 0, 0)) %>% 
  ggplot(aes(factor(age), per * 100, fill = school)) + 
    geom_bar(stat='identity', position = "dodge")  + 
    xlab("Age") +
    ylab("Relative frequency in %") +
    ggtitle("Bar chart for the age by schools") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank())
```
In the relative frequency plot, it is clear that students in **GP** are generally younger that in **MS**. In addition, age distribution in **MS** is centered that age group 18 is an absolute majority while age distribution in **GP** is spread that no age group accounts for more than thirty percent.

### Family Education Level

In this section, we will analyze the students' family education level in the two schools. Since the dataset offers the education level of both parents, we will use the higher one to indicate the family education level of a student, i.e. `max(Medu, Fedu)`.

```{r}
HH::likert(Pedu~., 
           stu %>%  mutate(Pedu = pmax(Medu, Fedu)) %>%
             count(school, Pedu, name = "Freq") %>% 
             pivot_wider(names_from = school, 
                         values_from = Freq) %>% 
             mutate(GP = 100 * GP / sum(GP),
                    MS = 100 * MS / sum(MS)),
           positive.order = TRUE,
           main = "Family education level in two schools",
           xlab = "percent", ylab = "")
```
According to the plot above, we can find that students' failmy education level in **GP** is generally higher that in **MS**.

### Sex & Address

In this section, we are going to look at the students' distribution interms of their gender and home address type (Urban or Rural).

```{r}
HH::likert(grp~., 
           union(stu %>% 
                   count(school, sex) %>% 
                   pivot_wider(names_from = school, values_from = n) %>% 
                   rename(grp = sex) %>% 
                   mutate(GP = 100 * GP / sum(GP),
                          MS = 100 * MS / sum(MS)),
                 stu %>% 
                   count(school, address) %>% 
                   pivot_wider(names_from = school, values_from = n) %>% 
                   rename(grp = address) %>% 
                   mutate(GP = 100 * GP / sum(GP),
                          MS = 100 * MS / sum(MS))),
           positive.order = TRUE,
           main = "% of gender and address in two schools",
           xlab = "percent", ylab = "")
```

In terms of gender, there is nearly no difference between two schools. Both **GP** and **MS** have a **half-half** gender distribution. However, only 20% students in **GP** are from rural area while there is no significance difference in students' address type in **MS**.

### Reason of Chosing School

In this section, we are going to multivariate analysis in terms of the association between schools, reasons, and family education level.

#### Reason & School

According to the plot below, there is association between schools and reasons. The **Pearson's Chi-squared** test rejects the hypothesis that two variables are not relevant with a p-value of 0.006, which is consistent with our observation.

Consequently, we can obtain that **GP** has a better reputation in terms of reason of choice while course offered by **MS** is more attractive.

```{r}
mosaic(reason ~ school, 
       stu %>% count(school, reason, name = "Freq"), 
       direction = c("v", "h"))
```

#### Reason Adr

According to the plot below, rural students are more likely to choose **MS** other than **GP** due to `home` and `other` reasons. A possible explanation would be **MS**'s study schedule is flexible. On the other hand, urban students tend to choose **MS** more often than **GP** because of `course`.

```{r}
mosaic(reason ~ address + school, 
       stu %>% count(school, reason, address, name = "Freq"), 
       direction = c("v", "v", "h"))
```
